---
title: "A tutorial for TMscoreAlign"
author: "Kevin Zhu"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{A tutorial for TMscoreAlign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup1, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5.5,
               dev.args=list(pointsize=10),
               par = TRUE, # needed for setting hook 
               collapse = TRUE, # collapse input & ouput code in chunks
               warning = FALSE)
knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(family = "sans", mar=c(4.1,4.1,1.1,1.1), mgp=c(3,1,0), tcl=-0.5)
})
set.seed(1) # for exact reproducibility
```

```{r setup}
require("devtools")
devtools::install_github("kevqyzhu/TMscoreAlign")
library(TMscoreAlign)
```

## Introduction
`TMscoreAlign` is an R package that implements the TM-align program in R 
([Zhang & Skolnick, 2005](https://doi.org/10.1002/prot.20264)). 
TM-align is a method for aligning proteins based on TM-score 
(Template Modelling score), which calculates topological similarity between two 
protein structures 
([Zhang & Skolnick, 2004](https://doi.org/10.1093/nar/gki524)). TM-score 
improves upon existing means of structural comparison such as RMSD 
(root-mean-square deviation) as it is length-independent and more sensitive to 
global similarities between structures rather than local deviations. This 
package is targeted for structural biologists who may use this package to 
investigate different conformations of the same protein, informing a structural
basis of protein functions. The tool includes functions that performs structure 
alignment, calculates TM-scores and RMSD, and visualizes aligned structures. 

See `help(package = "TMscoreAlign")` for further details. 
To download **TMscoreAlign**, use the following commands:

``` r
require("devtools")
devtools::install_github("kevqyzhu/TMscoreAlign", build_vignettes = TRUE)
library("TMscoreAlign")
```

To list all sample functions available in the package:
``` r
ls("package:TMscoreAlign")
```

To list all sample datasets available in the package:
``` r
data(package = "TMscoreAlign")
```

## Workflow

### Step 1: Preparing the inputs 

First, we will read in two PDB files. These will contain the protein structures
that we will perform structural alignment on. Note that PDB files can have 
multiple chains, so it is necessary to also specify for each PDB file which 
chain we want to use in this analysis.

``` r
# our first PDB file
pdb_file1 <- system.file("extdata", "1LNIA_decoy1_4.pdb", 
                          package="TMscoreAlign"
                          )
# our second PDB file
pdb_file2 <- system.file("extdata", "1LNIA_decoy2_180.pdb", 
                          package="TMscoreAlign"
                          )

# the chain identifier for pdb_file1
chain1 <- 'A'
# the chain identifier for pdb_file2
chain2 <- 'A'
```

### Step 2: Read and align
Now that we have specified the paths for our PDB files, as well as the chain
identifiers for each file, we can actually begin aligning them. The 
`get_alignment` function takes as input these file paths and chain identifiers.

Note that before structural alignment occurs, we must first identify which 
residues appear in both the template and target structures. Thus, when comparing
two protein structures that have the same residue order, we can use the 
C-alpha (the carbon that is one away from an aldehyde or ketone group) ordering.
In this case, we input "index" into the `method` parameter. When comparing two 
protein structures that have different residue orderings, a sequence alignment 
can be performed first by inputting "alignment" into the `method` parameter.

``` r
# aligning residues by index before the structural alignment
alignment <- get_alignment(pdb_file1, pdb_file2, 
                          chain1, chain2, method="index",
                          optimize = FALSE
                          )
                          
# using a sequence alignment before the structural alignment
alignment <- get_alignment(pdb_file1, pdb_file2, 
                          chain1, chain2, method="alignment",
                          optimize = FALSE
                          )
```

The output of this function is a `List` that contains structure alignment 
results, including alignment parameters, coordinates, and other information.
The list has the following elements:

* `N`: Numeric. Number of common residues in the alignment.
* `coord1`: Matrix. 3D coordinates of C-alpha atoms for common residues in the 
first structure.
* `coord2`: Matrix. 3D coordinates of C-alpha atoms for common residues in the 
second structure.
* `values`: Numeric vector. Initial alignment parameters.

Note that in the `get_alignment` call, we specified `optimize = FALSE`. This 
initializes the default alignment parameters into `values` but does not optimize
them.

``` r
# we can instantiate the alignment parameters and then optimize them directly 
alignment <- get_alignment(pdb_file1, pdb_file2, 
                          chain1, chain2, method="alignment",
                          optimize = TRUE
                          )
```

### Step 3: Optimize alignment

Now that we have our alignment parameters, we can optimize them. The 
`optimize_alignment` function takes as input the values we got from 
`get_alignment` and returns the parameters that minimize the objective function 
(TM-score). This makes use of the limited-memory 
Broyden–Fletcher–Goldfarb–Shanno algorithm from the `stats::optim`function.

``` r
alignment <- optimize_alignment(alignment)

```

We can also change the maximum number of iterations for optimization. This 
optimization step can be initialized with default values or continue from a 
given set of values using the `restart` parameter

``` r
alignment <- optimize_alignment(alignment, maxit = 400, restart = TRUE)

```

### Step 4: Getting alignment metrics and visualizing results

With the alignment parameters, we can get the TM-score, local TM-scores, and RMSD 
values of the alignment.

``` r
get_tmscore(alignment)
get_tm_samples(alignment)
get_rmsd(alignment)

```

We can also write a PDB file using the alignment parameters. The resulting file
contains two chains that represent the aligned protein structures. This output
PDB file can subsequently be visualized.

``` r
write_pdb(alignment, outputfile = "out.pdb", appended = TRUE, 
          pdb_file1, pdb_file2, chain1, chain2
          )
visualize_alignment_pdb("out.pdb")
```

## Package References

Zhu, K. (2023) TMscoreAlign: Protein structure alignment with TM-score.
  Unpublished. URL https://github.com/kevqyzhu/TMscoreAlign.

## Other References

Berman, H., Henrick, K. and Nakamura, H. Announcing the worldwide Protein Data
Bank. *Nature Structural & Molecular Biology*, 10, 980 (2003). 
https://doi.org/10.1038/nsb1203-980

Borchers H (2022). _pracma: Practical Numerical Math Functions_. R package 
version 2.4.2 https://CRAN.R-project.org/package=pracma

Byrd, R. H., Lu, P., Nocedal, J., & Zhu, C. (1995). A limited memory 
algorithm for bound constrained optimization.
*SIAM Journal on Scientific Computing*, 16(5), 1190–1208. 
https://epubs.siam.org/doi/10.1137/0916069

Deng, H., Jia, Y., and Zhang, Y. (2016). 3DRobot: Automated Generation of 
Diverse and Well-packed Protein Structure Decoys. *Boinformatics*, 32, 378-387.
https://doi.org/10.1093/bioinformatics/btv601

Grant, B.J. et al. (2006). _Bio3D: An R package for the comparative analysis
of protein structures_. *Bioinformatics*, 22, 2695--2696. 
https://doi.org/10.1093/bioinformatics/btl461
R package version 2.4.4, 
https://cran.r-project.org/web/packages/bio3d/index.html

Jordi Cruzado (https://math.stackexchange.com/users/96872/jordi-cruzado),
Explain 3d transformation matrix..., URL (version: 2023-05-30):
https://math.stackexchange.com/q/532974

Jur van den Berg 
(https://math.stackexchange.com/users/91768/jur-van-den-berg), Calculate 
Rotation Matrix to align Vector $A$ to Vector $B$ in $3D$?, URL 
(version: 2016-09-01): https://math.stackexchange.com/q/476311

Pagès H, Aboyoun P, Gentleman R, DebRoy S (2023).
_Biostrings: Efficient manipulation of biological strings_. 
R package version 2.68.1, https://doi.org/10.18129/B9.bioc.Biostrings

R Core Team (2023). R: A language and environment for statistical computing. 
R Foundation for Statistical Computing, Vienna, Austria. 
https://www.R-project.org/

Su W, Johnston B (2021). 
_r3dmol: Create Interactive 3D Visualizations of Molecular Data_. 
R package version 0.1.2, https://CRAN.R-project.org/package=r3dmol

Zhang, Y., and Skolnick, J. (2004). Scoring function for automated assessment 
of protein structure template quality. 
*Proteins, Structure, Function, and Bioinformatics*, 57(4), 702–710. 
https://doi.org/10.1002/prot.20264

----

```{r sessioninfo, message=FALSE}
sessionInfo()
```
